Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures

### 2020-04-20 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
removed for loops for efficiency, but did not fix pca convergence again.

---
title: Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multiple imputation by chained equations (mice) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Missing data problems are ubiquitous and pervasive in the social and behavioral sciences. If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result. Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset. With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed. The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations. By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness. Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18]. A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: missingness percentage = proportion of cases with one or more missing values **
<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

@@

From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$. $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:

@@

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic. **Added later: High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**

@@

Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R]. The simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy). 
<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->

>>>>

---
title: Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multiple imputation by chained equations (MICE) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Missing data problems are ubiquitous and pervasive in the social and behavioral sciences. If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result. Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset. With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed. The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations **see figure 1??**. By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness. Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18]. A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1].  **
<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

@@

From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why**. $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:

@@

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$ **explain that the randomness induced by the MICE algorithm would take care of this??**. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic. **Added later: High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**

@@

**Add to this section: 1) explicit mention of simulation conditions; 2) emphasize that the plots are averages across repetitions, not within MICE; 3) sample effects due to single complete dataset** Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R]. The simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy). 
<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->
